R Notebook

### 2021-07-14 draft implementation 2.Rmd
change in section

In this section we present a practical illustration of the algorithms discussed. For the sake of simplicity we use a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example will be used also to illustrate more involving filtering strategies described in this work. We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models. \
The filtering strategy is applied to 50 simulated data. Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}

@@

\
\textbf{Implementation}
Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm. The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm. We indicate with $n$ the sample size of time observations and with $N$ the generated sample size for each step of the Sequenial Monte Carlo. 
\begin{algorithm} SIS filter for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{SISfun} function implemented in \R replicate this steps.
\
\hrule
\hrule
\code{SISfun(data,N,m0,C0,tau,sigma)}
\hrule
\textbf{Arguments}

@@

\textbf{Implementation}
In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX. The overall strategy replicate the SIS filter with the addition of a ESS-based resampling step that applies when the effective sample size is smaller than a predetermined threshold opportunely chosen (in our example we decide to follow a common rule of thumb consisting in setting the threshold at $N/2$). The steps are presented in the following algorithm.
\begin{algorithm} BPF for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
These steps are resumed in our \code{PFfun} function.
\
\hrule
\hrule
\code{PFfun(data,N,m0,C0,tau,sigma,r)}
\hrule
\textbf{Arguments}

@@

ggarrange(p1,c1,p2,c2,p3,c3,nrow=3,ncol=2,labels=c("N=100","","N=1000","","N=10000",""))
```
\
\textbf{Implementation}
\
The Bootstrap Particle Filter Approach for the Random Walk plus Noise Model described in Section XX can be improved accounting for the observations in the importance transition density and it consists of generating $x_{t}$ from its conditional distribution given $x_{t-1}$ and $y_{t}$. In the Normal model we are considering, the optimal proposal will be a Normal density as well with mean and variance given by 
\begin{align*}
\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
\end{align*}
On the other hand, the incremental weights, using this importance transition density, are proportional to the conditional density of $y_{t}$ given $x_{t-1}=x_{t-1}^{(i)}$ , i.e $N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})$, evaluated at $y_{t}$. In other words the algorithm implemented in \R is:
\begin{algorithm} GPF for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item Compute $\sigma_{opt}^{2}$ 
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute $\mu_{opt}$
\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
The \code{GPFfun} function resume this passages.
\
\hrule
\hrule
\code{GPFfun(data,N,m0,C0,tau,sigma,r)}
\hrule
\textbf{Arguments}

>>>>

In this section we present a practical illustration of the algorithms discussed. For the sake of simplicity we use a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example will be used also to illustrate more involved filtering strategies described in this work. We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models. \
The filtering strategy is applied to 50 simulated data. Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}

@@

\
\section{Implementation}
Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm. The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm. We indicate with $n$ the sample size of time observations and with $N$ the generated sample size for each step of the Sequenial Monte Carlo. 
\begin{algorithm} SIS filter for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{SISfun} function implemented in \R replicate this steps.
\
\hrule
\hrule
\code{SISfun(data,N,m0,C0,tau,sigma)}
\hrule
\textbf{Arguments}

@@

\section{Implementation}
In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX. The overall strategy replicate the SIS filter with the addition of a ESS-based resampling step that applies when the effective sample size is smaller than a predetermined threshold opportunely chosen (in our example we decide to follow a common rule of thumb consisting in setting the threshold at $N/2$). The steps are presented in the following algorithm.
\begin{algorithm} BPF for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
These steps are resumed in our \code{PFfun} function.
\
\hrule
\hrule
\code{PFfun(data,N,m0,C0,tau,sigma,r)}
\hrule
\textbf{Arguments}

@@

ggarrange(p1,c1,p2,c2,p3,c3,nrow=3,ncol=2,labels=c("N=100","","N=1000","","N=10000",""))
```
\
\section{Implementation}
\
The Bootstrap Particle Filter Approach for the Random Walk plus Noise Model described in Section XX can be improved accounting for the observations in the importance transition density and it consists of generating $x_{t}$ from its conditional distribution given $x_{t-1}$ and $y_{t}$. In the Normal model we are considering, the optimal proposal will be a Normal density as well with mean and variance given by 
\begin{align*}
\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
\end{align*}
On the other hand, the incremental weights, using this importance transition density, are proportional to the conditional density of $y_{t}$ given $x_{t-1}=x_{t-1}^{(i)}$ , i.e $N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})$, evaluated at $y_{t}$. In other words the algorithm implemented in \R is:
\begin{algorithm} GPF for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item Compute $\sigma_{opt}^{2}$ 
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute $\mu_{opt}$
\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{enumerate}
\end{itemize}
\end{algorithm}
The \code{GPFfun} function resume this passages.
\
\hrule
\hrule
\code{GPFfun(data,N,m0,C0,tau,sigma,r)}
\hrule
\textbf{Arguments}
